---
title: "STAT462 Assignment 2"
author: "Thomas Waldin - 17775654"
date: "2025-08-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The objective of this report is to investigate datasets and demonstrate understanding of
classification algorithms.

# Part A: Pumpkin Seeds
A logistic regression was trained for automatic classification of two types 
of pumpkin seeds, Çerçevelik and Ürgüp Sivrisi, based on geometric (measured) features of these 
seeds.

### 1)
The datasets seeds_training.csv and seeds_test.csv were loaded. Since the class labels (the 
seed types) contain characters the class features were encoded as a factor 
variable in both training and test set. 

```{r, results='hide', warning=FALSE, message=FALSE}
# Import libraries
library(tidyverse)
library(caret)
library(MASS)
library(ggplot2)

# Read data
df_train <- read_csv("seeds_training.csv")
df_test <- read_csv("seeds_test.csv")


# Encode Class variable as factor
seeds_train <- df_train %>%
  mutate(Class = as.factor(Class))

seeds_test <- df_test %>%
  mutate(Class = as.factor(Class))

```

### 2) 
A multiple logistic regression classifier was fit to the data using all of the features.
The accuracy on the test set was found to be 0.888.

```{r}
# Fit a multiple logistic regression using all features
lm <- glm(Class ~ ., 
          data = seeds_test, 
          family = binomial)

# Make predictions on test set
pred_prob <- predict(lm,
                       newdata = seeds_test,
                       type = "response")

# Convert probabilities to class predictions (using 0.5 as threshold)
class_labels <- levels(as.factor(seeds_test$Class))
pred_class <- ifelse(pred_prob > 0.5, class_labels[2], class_labels[1])

# Calculate and print accuracy
accuracy <- mean(pred_class == seeds_test$Class)
cat("Accuracy on test set:", round(accuracy, 4), "\n")
```

### 3) 
A confusion matrix for the model was generated based on its predictions on the test set.
The concept of sensitivity and specificity do not make a lot of sense in this example. Neither seed is inherently positive or negative, each are just different varieties, and the choice of which seed being positive is arbitrary. Without further context it is unknown if one misclassification is more problematic than the other, and it is likley that it is equally important to classify each correctly. The accuracy or the balanced accuracy is likely a better metric in this case.



```{r}
# Compute confusion matrix
confusionMatrix(as.factor(pred_class),as.factor(seeds_test$Class), positive = "Cercevelik")

```

### 4)
Observing the summary of the GLM model, it appeared that the area, the convex area, and the equivilant diameter of the seeds were the most significant features.

```{r}
summary(lm)
```

### 5) 
Another GLM model was trained, only using the identified three most significant features. The accuracy and confusion matrix were generated so that they could be compared with the previously trained model. The accuracy on the (arbitrary) positive class was much lower in the second model (0.572 vs. 0.888) and the balanced accuracy was also much lower (0.576 vs. 0.886). As previously mentioned, the balanced accuracy would be a good metric to consider in this case, so the model with all the features should be selected. There is likely a model between these two, with more of the significant features, but no all features which may outperform both models that were considered.

```{r}
# Fit a multiple logistic regression using select features
lm <- glm(Class ~ Area + Convex_Area + Equiv_Diameter, 
          data = seeds_test, 
          family = binomial)

# Make predictions on test set
pred_prob <- predict(lm,
                       newdata = seeds_test,
                       type = "response")

# Convert probabilities to class predictions (using 0.5 as threshold)
class_labels <- levels(as.factor(seeds_test$Class))
pred_class <- ifelse(pred_prob > 0.5, class_labels[2], class_labels[1])

# Calculate and print accuracy
accuracy <- mean(pred_class == seeds_test$Class)
cat("Accuracy on test set:", round(accuracy, 4), "\n", "\n")

# Compute confusion matrix
confusionMatrix(as.factor(pred_class),as.factor(seeds_test$Class), positive = "Cercevelik")
```

# Part B: Colours
 Colours can be coded via their RGB, three integers between 0 and 255.
In this exercise, RGB values were mapped to colour names, but only considering the colour space where G = 0.

### 1)
The colors_train.csv dataset was loaded and inspected, it was found that there are five classes in the dataset - red, pink, purple, brown, and blue.

```{r, results='hide'}
# Load the dataset colors_train
df_train <- read_csv('colors_train.csv')
df_test <- read_csv('colors_test.csv')

# Encode Class variable as factor
colors_train <- df_train %>%
  mutate(color = as.factor(color))
colors_test <- df_test %>%
  mutate(color = as.factor(color))
```

```{r}
# Examine structure of dataframe
str(colors_test)
```

### 2)
A QDA algorithm was fit to the classification problem and a confusion matrix was generated.

```{r}
# Fit a QDA algorithm
qda.fit <- qda(color ~ r + b, data = colors_train)
pred.qda.fit = predict(qda.fit, newdata = colors_test)$class

# Compute confusion matrix
confusionMatrix(as.factor(pred.qda.fit),as.factor(colors_test$color), positive = "Yes")
```

### 3)
Figure 1 visualises the decision boudaries generated by the QDA algorithm.

```{r, echo=FALSE}

# Create a grid of points to evaluate the decision boundary
x_range <- range(colors_train$r)
y_range <- range(colors_train$b)
grid_points <- expand.grid(
  r = seq(x_range[1], x_range[2], length.out = 100),
  b = seq(y_range[1], y_range[2], length.out = 100)
)

# Predict on the grid
grid_predictions <- predict(qda.fit, newdata = grid_points)$class

# Combine grid with predictions
grid_df <- data.frame(grid_points, predicted_class = grid_predictions)

# Plot with color-appropriate styling
ggplot() +
  # Decision boundary regions
  geom_tile(data = grid_df, aes(x = r, y = b, fill = predicted_class), alpha = 0.3) +
  # Actual data points
  geom_point(data = colors_test, aes(x = r, y = b, color = color), size = 2) +
  
  # Color scales that match actual colors
  scale_fill_manual(
    name = "Predicted Color",
    values = c(
      "red" = "lightcoral",
      "pink" = "lightpink", 
      "brown" = "tan",
      "purple" = "plum",
      "blue" = "lightblue"
    )
  ) +
  scale_color_manual(
    name = "Actual Color",
    values = c(
      "red" = "red",
      "pink" = "hotpink",
      "brown" = "brown", 
      "purple" = "purple",
      "blue" = "blue"
    )
  ) +
  
  # Labels appropriate for RGB color space
  labs(
    x = "Red Value (0-255)", 
    y = "Blue Value (0-255)", 
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

```

###### Figure 1: Classification boundaries as defiend by the QDA algorithm. The test data and with actual classifications are plotted over the boundaries as points.

### 4)
The algorithm was tested with the point (200, 0, 200). The algorithm calls this colour 'pink', although the predicted probability of 'purple' is high, suggesting it may lay close to the boundary, which was confirmed by examining Figure 1.

```{r}
# Create a data frame with the test point
test_point <- data.frame(r = 200, b = 200)

# Predict the color class
predicted_color <- predict(qda.fit, newdata = test_point)

# Get the predicted class
predicted_color$class

# Get the prediction probabilities for all classes
predicted_color$posterior
```



